Austraits -- a curated plant trait database for the Australian flora

### 2020-07-09 ms.Rmd
Update data source to v 2.0

austraits <- readRDS("data/austraits.rds")
```

@@

to_exclude <- c("Barlow_1981","BRAIN_2007","CPRR_2002","Kew_2019_1","Kew_2019_2","Kew_2019_3",
"Kew_2019_4","Kew_2019_5","Kew_2019_6","NHNSW_2016","NTH_2014","RBGK_2014","RBGSYD_0000","RBGSYD_2014","SAH_2014", "Wheeler_2002", "Hyland_2003", "SAH_2014")

data_contributors <- austraits$contributors %>% select(-role) %>% 
    # remove authors 
    filter(!name %in% c("Peter Clarke", "Barbara Rice", "Westoby Lab")) %>% 
    # intstituional collections
    filter(!dataset_id %in% to_exclude) %>%
    select(-dataset_id) %>% 
    distinct() %>% 
    mutate(surname = str_split(name, " ") %>% map_chr(last)) %>% 
    arrange(surname) %>% rename(affiliation = institution)

# Need to check how these institutions and individuals are to be recognised
# austraits$contributors %>% select(-role) %>% 
#     filter(dataset_id %in% to_exclude) %>% 
#     filter(name!="Rachael Gallagher") %>% 
#   write_csv("authors_check.csv")

authors <- bind_rows(authors_build, 
                     data_contributors %>% filter(! (name %in% authors_build$name)))

@@

We introduce the AusTraits database - a compilation of measurements of plant traits for species in the Australian Flora (hereafter AusTraits). AusTraits synthesises data on `r austraits$traits$trait_name %>% n_distinct()` traits across `r austraits$traits$species_name %>% n_distinct()` species from field campaigns, published literature, taxonomic monographs, and individual species descriptions. Traits
vary in scope from physiological measures of performance (e.g. photosynthetic gas exchange, water-use efficiency) to morphological
parameters (e.g. leaf area, seed mass, plant height) which link to aspects of ecological variation. AusTraits contains curated and
harmonised individual-, species- and genus-level observations coupled to, where available, contextual information on site
properties. This data descriptor provides information on Version `r austraits$build_info$version` of AusTraits which contains data for `r austraits$traits %>% nrow()` trait-by-species combinations. We envision AusTraits as an on-
going collaborative initiative for easily archiving and sharing trait data to increase our collective understanding of the Australian flora.

@@

Here we introduce the AusTraits database (hereafter simply AusTraits) - a compilation of plant traits for the Australian flora. Currently, AusTraits draws together `r austraits$sources %>% length()` primary sources – including published scientific papers, floras, and taxonomic species descriptions– and contains `r austraits$traits %>% nrow()` measurements spread across `r austraits$traits$trait_name %>% n_distinct()` different traits for  `r austraits$traits$species_name %>% n_distinct()` species (Table 1; available online only). By providing a harmonised and curated dataset on `r austraits$traits$trait_name %>% n_distinct()` plant traits, AusTraits contributes substantially to filling the gap in Australian and global biodiversity resources. Prior to the development of AusTraits, data on Australian plant traits existed largely as a series of disconnected datasets collected by individual laboratories or initiatives (but see @guerin2017opportunities). Global initiatives, such as the TRY database [@Kattge-2020] and Open Traits Network [@Gallagher-2019], have been working towards sharing of data about plant traits though large numbers of Australian plant species remain unrepresented.

To assemble AusTraits from diverse primary sources and make it available for reuse, we needed to overcome three main types of challenges (Figure 1). These included: 1) **Accessing** data from diverse original sources, including field studies, online databases, scientific articles, and published taxonomic floras; 2) **Harmonising** these diverse datasets into a federated resource, with common units, trait names, and data formats; and 3) **Distributing** versions of the data under suitable license. Successive versions AusTraits will iterate through these to incorporate new data and correct identified errors.  We developed a workflow for harmonising and distributing data, drawing from emerging community standards and our collective experience building traits databases. Combined, the tools applied ensure current best-practices in the construction and distribution of data resources, by achieving the following outcomes:

  - A reproducible and extensible compilation process from raw data to assembled dataset;
  - Transparent mapping of original values (species names, trait names, units) into the harmonised dataset;
  - High quality data for end users; 
  - Easy addition of new data and correction of community-reported errors; 
  - Satisfying the FAIR principles [@wilkinson2016fair], i.e. standards needed to make data resources Findable, Accessible, Interoperable, and Reusable; and
  - Recognition of contributors through citation of this resource and primary sources.

@@

  - **Automated tests** which confirm that the species names provided can be mapped onto accepted names on The Plant List, that values for continuous traits fall within the accepted range for the trait, and that values for categorical traits are on a list of accepted values maintained by the creators. Data that does not pass these tests is moved to a separate spreadsheet (“excluded_data”) that is also made available for use if desired.

@@

@@

```{r austraits}
## Assumes to items exist in global name space
austraits <- readRDS("data/austraits.rds")

@@

AusTraits harmonises data on `r austraits$traits$trait_name %>% unique() %>% length()` traits from `r austraits$methods$dataset_id %>% unique() %>% length()` different sources, including field campaigns, published literature, taxonomic monographs, and individual species descriptions.

@@

![**Figure 2.** (a.) Locations where the trait data were collected and the corresponding number of species. (b.) AusTraits’ sites (grey) within Australia’s precipitation-temperature space (black) superimposed upon Whittaker’s major biomes diagram. Data from WorldClim (worldclim.org). (c.) Overview of the number of species with which traits are measured within each major plant organ and trait category.](figures/austraits_fig2_blue.png)

![**Figure 3.** The phylogenetic representation of traits in the AusTraits database for the subset of 2000 randomly sampled species. The heatmap colour intensities denote the number of traits measured within each major plant organ. The most widespread family names (with more than ten species) are labelled on the edge of the tree.](figures/austraits_phlogenetic_coverage.png)

@@

data_geo <- 
  left_join(by=c("dataset_id", "site_name"),
      select(austraits$traits, -original_name), sites) %>%
  left_join(by="species_name",
      select(austraits$taxonomy, species_name, family)) %>% 
  left_join(sources_all, by=c("dataset_id"))

n_records <- 
  data_geo %>% 
  group_by(trait_name) %>%
  summarise(
            Description = "",
            Type= "",
            `Records (all)` = n(),
            `Records (geo.)` = sum(!is.na(`longitude (deg)`) & !is.na(`latitude (deg)`)),
            Studies = n_distinct(dataset_id),
            Species = n_distinct(species_name),
            Families = n_distinct(family),
            `Data Citations` = refs %>% unique() %>% paste(collapse = "; ")
            ) %>% 
  left_join(trait_categories, by=c("trait_name")) %>% 
  select(Tissue = tissue, Category = category, Trait = trait_name, everything()) %>% 
  arrange(Tissue, Category, Trait)

@@

```{r, results="asis"}
print_defintions_element <- function(elements) {
  if(elements$type == "character") {
    sprintf("**Content:** %s\n", elements$value) %>% 
      writeLines()
  }

>>>>

austraits <- readRDS("data/austraits_2.0.0.rds")
```

@@

data_contributors <- 
  austraits$contributors %>% 
    filter(
          # old studies
          !dataset_id %in% c("Lawes_2012", "Westman_1977", "Hall_1981", "Hocking_1982", "Hocking_1986", "Toelken_1996", "Brock_1993", "Forster_1992", "Forster_1995", "Goble_1981"),
           !role %in% c("assistant"),
           # deceased
           !name %in% c("Peter Clarke", "Barbara Rice", "Peter Myerscough", "Lyn Craven", "Harold Trevor Clifford", "William Cooper")
           ) %>% 
    mutate(year = dataset_id %>% str_split_fixed("_", 4) %>% .[,2]) %>% 
    arrange(year) %>%
    # take most recent address for each author
    group_by(name) %>% 
    summarise(institution = last(institution),
              dataset_ids = paste(dataset_id, collapse = ", ")) %>% 
    ungroup() %>% 
    distinct() %>% 
    mutate(surname = str_split(name, " ") %>% map_chr(last)) %>% 
    rename(affiliation = institution) %>%
#    bind_rows(
#      read_csv("data/data_pending.csv")) %>%
    arrange(surname) %>% 
    select(-surname)

authors <- bind_rows(authors_build, 
                     data_contributors %>% filter(! (name %in% authors_build$name))
                    )

#  authors %>% select(dataset_ids, everything()) %>% mutate_all(~ replace_na(.x, "")) %>% write_csv("data/author_list.csv")

# Update author info from google sheets

@@

We introduce the AusTraits database - a compilation of measurements of plant traits for taxa in the Australian Flora (hereafter AusTraits). AusTraits synthesises data on `r austraits$traits$trait_name %>% n_distinct()` traits across `r austraits$traits$taxon_name %>% n_distinct()` taxa from field campaigns, published literature, taxonomic monographs, and individual taxa descriptions. Traits
vary in scope from physiological measures of performance (e.g. photosynthetic gas exchange, water-use efficiency) to morphological
parameters (e.g. leaf area, seed mass, plant height) which link to aspects of ecological variation. AusTraits contains curated and
harmonised individual-, species- and genus-level observations coupled to, where available, contextual information on site
properties. This data descriptor provides information on Version `r austraits$build_info$version` of AusTraits which contains data for `r austraits$traits %>% nrow()` trait-by-taxa combinations. We envision AusTraits as an on-
going collaborative initiative for easily archiving and sharing trait data to increase our collective understanding of the Australian flora.

## Todo

Refs

* Australian Plant Census, Centre of Australian National Biodiversity Research, Council of Heads of Australasian Herbaria, (14/05/2020) https://id.biodiversity.org.au/tree/51354547
* Australian Plant Name Index (continuously updated), Centre of Australian National Biodiversity Research, www.biodiversity.org.au/nsl/services/apni (14/05/2020)

@@

Here we introduce the AusTraits database (hereafter simply AusTraits) - a compilation of plant traits for the Australian flora. Currently, AusTraits draws together `r austraits$sources %>% length()` primary sources – including published scientific papers, floras, and taxonomic descriptions– and contains `r austraits$traits %>% nrow()` measurements spread across `r austraits$traits$trait_name %>% n_distinct()` different traits for  `r austraits$traits$taxon_name %>% n_distinct()` taxa (Table 1; available online only). By providing a harmonised and curated dataset on `r austraits$traits$trait_name %>% n_distinct()` plant traits, AusTraits contributes substantially to filling the gap in Australian and global biodiversity resources. Prior to the development of AusTraits, data on Australian plant traits existed largely as a series of disconnected datasets collected by individual laboratories or initiatives (but see @guerin2017opportunities). Global initiatives, such as the TRY database [@Kattge-2020] and Open Traits Network [@Gallagher-2019], have been working towards sharing of data about plant traits though large numbers of Australian plant taxa remain unrepresented.

To assemble AusTraits from diverse primary sources and make it available for reuse, we needed to overcome three main types of challenges (Figure 1). These included: 1) **Accessing** data from diverse original sources, including field studies, online databases, scientific articles, and published taxonomic floras; 2) **Harmonising** these diverse datasets into a federated resource, with common units, trait names, and data formats; and 3) **Distributing** versions of the data under suitable license. Successive versions of AusTraits will iterate through these to incorporate new data and correct identified errors.  We developed a workflow for harmonising and distributing data, drawing from emerging community standards and our collective experience building traits databases. Combined, the tools applied ensure current best-practices in the construction and distribution of data resources, by achieving the following outcomes:

  - A reproducible and extensible compilation process from raw data to assembled dataset;
  - Transparent mapping of original values (taxon names, trait names, units) into the harmonised dataset;
  - High quality data for end users; 
  - Easy addition of new data and correction of community-reported errors; 
  - Satisfying the FAIR principles [@wilkinson2016fair], i.e. standards needed to make data resources Findable, Accessible, Interoperable, and Reusable; and
  - Recognition of contributors through citation of this resource and primary sources.

@@

  - **Automated tests** which confirm that the taxa names provided can be mapped onto accepted names on The Plant List, that values for continuous traits fall within the accepted range for the trait, and that values for categorical traits are on a list of accepted values maintained by the creators. Data that does not pass these tests is moved to a separate spreadsheet (“excluded_data”) that is also made available for use if desired.

@@

We achknowelde XXXXXXXXXXx institutions, who

@@

```{r austraits}
## Assumes to items exist in global name space
austraits <- readRDS("data/austraits_2.0.0.rds")

@@

AusTraits harmonises data on `r austraits$traits$trait_name %>% unique() %>% length()` traits from `r austraits$methods$dataset_id %>% unique() %>% length()` different sources, including field campaigns, published literature, taxonomic monographs, and individual taxa descriptions.

@@

![**Figure 2.** (a.) Locations where the trait data were collected and the corresponding number of taxa. (b.) AusTraits’ sites (grey) within Australia’s precipitation-temperature space (black) superimposed upon Whittaker’s major biomes diagram. Data from WorldClim (worldclim.org). (c.) Overview of the number of taxa with which traits are measured within each major plant organ and trait category.](figures/austraits_fig2_blue.png)

![**Figure 3.** The phylogenetic representation of traits in the AusTraits database for the subset of 2000 randomly sampled taxa. The heatmap colour intensities denote the number of traits measured within each major plant organ. The most widespread family names (with more than ten taxa) are labelled on the edge of the tree.](figures/austraits_phlogenetic_coverage.png)

@@

data_geo <- 
  left_join(by=c("dataset_id", "site_name"),
      select(austraits$traits, -original_name), sites) %>%
  left_join(by="taxon_name",
      select(austraits$taxa, taxon_name, family)) %>% 
  left_join(sources_all, by=c("dataset_id"))

n_records <- 
  data_geo %>% 
  group_by(trait_name) %>%
  summarise(
            Description = "",
            Type= "",
            `Records (all)` = n(),
            `Records (geo.)` = sum(!is.na(`longitude (deg)`) & !is.na(`latitude (deg)`)),
            Studies = n_distinct(dataset_id),
            Taxon = n_distinct(taxon_name),
            Families = n_distinct(family),
            `Data Citations` = refs %>% unique() %>% paste(collapse = "; ")
            ) %>% 
  left_join(trait_categories, by=c("trait_name")) %>% 
  select(Tissue = tissue, Category = category, Trait = trait_name, everything()) %>% 
  arrange(Tissue, Category, Trait)

@@

```{r, results="asis", eval=FALSE}
print_defintions_element <- function(elements) {
  if(elements$type == "character") {
    sprintf("**Content:** %s\n", elements$value) %>% 
      writeLines()
  }
